\chapter{Types, Expressions, and Functions}
In this Chapter we will give a more systematic overview of \textsl{Haskell}.  In particular, we define
\begin{enumerate}
\item types, 
\item expressions, and
\item functions.
\end{enumerate}
At this point you might wonder why we don't also discuss statements and control structures.  The reason is
simple:  There are no statements in \textsl{Haskell}. Everything is an expression.  Neither are there control
structures. 

\textsl{Haskell} is a statically typed, purely functional programming language known for its expressive type
system and emphasis on immutability. In \textsl{Haskell}, types play a central role in the design and
implementation of programs. This section discusses the primitive types available in \textsl{Haskell}, their
properties, and how they form the building blocks for more complex types in the language. 

Primitive types in \textsl{Haskell} refer to the basic types that are built into the language and directly
supported by the compiler. These types include numbers, characters, booleans, and the unit type. Understanding
these types is crucial for both writing correct programs and for taking full advantage of \textsl{Haskell}'s
type system. 

\section{Primitive Types}
\textsl{Haskell} provides several predefined types, each serving different needs with respect to performance, precision,
and range. The main numeric types in \textsl{Haskell} include: 
\begin{enumerate}[(a)]
\item \texttt{Int} is a fixed-precision integer type, which means that its range is limited by the underlying
      hardware. Typically, \texttt{Int} is implemented as a 32-bit or 64-bit integer. Its limited range means that
      operations on very lar~ge numbers may result in overflow. 

      \paragraph{Example Usage:}
      \begin{lstlisting}[style=haskellstyle, language=Haskell]
 -- Defining an integer of type Int
 smallInt :: Int
 smallInt = 42
      \end{lstlisting}

\item \texttt{Integer} represents arbitrary-precision integers. This means that there is no fixed upper bound
      on the size of an \texttt{Integer} value, though operations may become slower as numbers grow larger. 

\paragraph{Example Usage:}
\begin{lstlisting}[style=haskellstyle, language=Haskell]
-- Defining an integer of type Integer
bigInt :: Integer
bigInt = 123456789012345678901234567890
\end{lstlisting}

\item \texttt{Float} is the type of single-precision floating-point numbers. While it may be faster and uses less
      memory, its precision is limited compared to \texttt{Double}. 

\paragraph{Example Usage:}
\begin{lstlisting}[style=haskellstyle, language=Haskell]
-- Defining a floating-point number of type Float
singlePrecision :: Float
singlePrecision = 3.14159
\end{lstlisting}

\item \texttt{Double} is a double-precision floating-point number, offering more precision at the cost of
      additional memory and potentially slower computation in some contexts. 

\paragraph{Example Usage:}
\begin{lstlisting}[style=haskellstyle, language=Haskell]
-- Defining a floating-point number of type Double
doublePrecision :: Double
doublePrecision = 2.718281828459045
\end{lstlisting}

\item \texttt{Char} is used to represent single Unicode characters. Characters in \textsl{Haskell} are enclosed
      in single quotes. 

\paragraph{Example Usage:}
\begin{lstlisting}[style=haskellstyle, language=Haskell]
-- A character literal
letterA :: Char
letterA = 'A'
\end{lstlisting}

\item \texttt{String} is used to represent Unicode strings.  In \textsl{Haskell}, strings are enclosed in double quotes.

\paragraph{Example Usage:}
\begin{lstlisting}[style=haskellstyle, language=Haskell]
-- A string literal
letterA :: Char
letterA = "Hello, World!"
\end{lstlisting}
In \textsl{Haskell} the type \texttt{String} is an alias for the type \texttt{[Char]}, which represents a list
of characters.

\item \texttt{Bool} represents boolean values. It has two possible values: \texttt{True} and \texttt{False}.

\paragraph{Example Usage:}
\begin{lstlisting}[style=haskellstyle, language=Haskell]
-- A boolean literal
isHaskellFun :: Bool
isHaskellFun = True
\end{lstlisting}

\item The \blue{unit type} is denoted by \texttt{()}.  This type has exactly one value, which is also written
  as \texttt{()}.  It is analogous to the concept of \textit{None} in \textsl{Python}. It is typically used
  when a function does not need to return any  meaningful value. 

\paragraph{Example Usage:}
\begin{lstlisting}[style=haskellstyle, language=Haskell]
-- A function that returns the unit type
printMessage :: String -> ()
printMessage msg = putStrLn msg
\end{lstlisting}
\end{enumerate}

\noindent
One of \textsl{Haskell}’s powerful features is its ability to perform type inference. For example, numeric literals in
\textsl{Haskell} are \blue{polymorphic}, i.e. they do not have a fixed type but rather a \blue{type class}.  This
concept will be discussed in a subsequent chapter.  For example, a literal such as \texttt{5} can be
interpreted as an \texttt{Int}, an \texttt{Integer}, a \texttt{Float}, or a \texttt{Double}, depending on the
context. This is achieved via type classes such as \texttt{Num}. 

\paragraph{Example:}
\begin{lstlisting}[style=haskellstyle, language=Haskell]
-- The literal 5 is polymorphic and can be any type that is an instance of Num.
polymorphicExample :: Num a => a
polymorphicExample = 5
\end{lstlisting}

\section{Composite Types: Lists and Tuples}

\textsl{Haskell} provides several composite types that allow for the grouping of values. Two of the most commonly used
composite types are \textbf{lists} and \textbf{tuples}. Both types enable the construction of complex data
structures by combining simpler types, yet they serve different purposes and have distinct characteristics. 

\subsection{Lists}
A \textbf{list} in \textsl{Haskell} is an \blue{homogeneous} ordered collection of elements.  The fact that a list is
homogeneous means that \colorbox{yellow}{all of elements must be of the same type.} Lists are one of the most
fundamental data structures in \textsl{Haskell} and are used extensively for 
processing sequences of data. Lists are denoted using square brackets, with elements separated by commas. For
example, the list containing the integers 1, 2, and 3 is written as: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
[1, 2, 3]
\end{lstlisting}
If \texttt{a} is any type, then the type of a list of type \texttt{a} is written as:
\[
[ a ]
\]
Lists are the workhorse of many functional languages and \textsl{Haskell} is no exception.  Therefore, \textsl{Haskell}
provides a rich set of functions for processing lists, such as \texttt{map}, \texttt{filter}, 
\texttt{foldr} and \texttt{foldl}.  These functions will be discussed later after we have discussed the syntax
of functions.

Lists are implemented as \href{https://en.wikipedia.org/wiki/Linked_list}{linked lists}, which means that
operations such as prepending an element (using the \texttt{:} operator) are very efficient. For example: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
-- Prepending 0 to an existing list:
numbers :: [Int]
numbers = 0 : [1, 2, 3]  -- results in [0, 1, 2, 3]
\end{lstlisting}
However, other operations have a linear complexity.  For example, finding the length of a list has a linear
complexity because the whole list needs to be traversed.  This is in contrast to the programming language
\textsl{Python}, where lists are implemented as dynamic arrays.  Hence, in \textsl{Python}, finding the length of a list
has complexity $\mathcal{O}(1)$.

\blue{Pattern matching} on lists is a powerful feature in \textsl{Haskell}. One common idiom is to match against the
empty list \texttt{[]} or a cons cell \texttt{(x:xs)}, where \texttt{x} is the head of the list and \texttt{xs}
is the tail. For example: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
sumList :: Num a => [a] -> a
sumList []     = 0
sumList (x:xs) = x + sumList xs
\end{lstlisting}
This recursive definition demonstrates how lists lend themselves naturally to inductive processing.

There is another very important difference between lists in \textsl{Python} and lists in \textsl{Haskell}: In
\textsl{Haskell}, lists are \blue{immutable}, i.e.~once we have constructed a list, there is no way to change
an element in this list.  This is similar to \textsl{tuples} in \textsl{Python}.

\subsection{Tuples}
In contrast to a list, a \textbf{tuple} is a composite type that can hold a fixed number of elements, which may
be of \colorbox{yellow}{different} types. Tuples are written using parentheses, with elements separated by commas. For instance, the tuple:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
("Alice", 30, True)
\end{lstlisting}
contains a \texttt{String}, an \texttt{Int}, and a \texttt{Bool}. The type of this tuple is written as:\\[0.2cm]
\hspace*{1.3cm}
\texttt{(String, Int, Bool)}
\\[0.2cm]
In general, the type of a tuple is denoted as:
\\[0.2cm]
\hspace*{1.3cm}
$(\texttt{t}_1, \texttt{t}_2, \cdots, \texttt{t}_n)$
\\[0.2cm]
Here, $\texttt{t}_i$ is the type of the $i^{\textrm{th}}$ component.

Unlike lists, tuples are \blue{heterogeneous}, i.e.~the elements can be of different types.  When declaring the
type of a tuple, the size, i.e.~the number of elements, is implicitly also defined.  Tuples are
\blue{immutable}.  In fact, every data structure in \textsl{Haskell} is immutable.  

Tuples are particularly useful when you need to group a set of values that
naturally belong together, such as coordinates or key-value pairs. 
Tuples support pattern matching, which allows functions to easily deconstruct them. For example, a function
that extracts the first element of a pair can be defined as: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
first :: (a, b) -> a
first (x, _) = x
\end{lstlisting}
Similarly, functions can be defined to operate on larger tuples by matching each component:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
describePerson :: (String, Int, Bool) -> String
describePerson (name, age, isEmployed) =
  name ++ " is " ++ show age ++ " years old and " ++
  (if isEmployed then "employed" else "unemployed") ++ "."
\end{lstlisting}
The previous example is easy to misunderstand because the function \texttt{describePerson} receives not three
elements but rather one element, which is a tuple of three elements.


\section{Haskell Expressions and Operators}
In \textsl{Haskell}, every construct is an expression that evaluates to a value. Unlike imperative languages where
statements perform actions, in \textsl{Haskell} even control constructs such as conditionals and pattern matching yield
results. One of the most powerful features of \textsl{Haskell} is its flexible and composable syntax for expressions,
which is largely governed by a rich set of infix operators. These operators come with fixed precedences and
associativities that determine the order in which parts of an expression are evaluated. In this section we
present a detailed discussion of \textsl{Haskell}’s operators, their precedence, associativity, and how they interact
within expressions. We will also provide a comprehensive table of many common operators, together with examples
to illustrate their usage.  

Before examining the operators, it is important to recall that \textsl{Haskell} function application (i.e., writing
\texttt{f x} to apply the function \texttt{f} to the argument \texttt{x}) has the highest precedence of all
operations. This means that in an expression like \texttt{f x + y}, the application of \texttt{f} to \texttt{x}
is performed first, and then the addition is carried out.

\subsection*{Operator Precedence and Associativity}
The precedence of an operator indicates how tightly it binds to its operands. Operators with higher precedence
are applied before operators with lower precedence. Associativity, on the other hand, determines how operators
of the same precedence are grouped in the absence of explicit parentheses. For example, left-associative
operators group from the left.  For example,
\\[0.2cm]
\hspace*{1.3cm}
\texttt{a - b - c}
\\[0.2cm]
is interpreted as
\\[0.2cm]
\hspace*{1.3cm}
\texttt{(a - b) - c}). 
\\[0.2cm]
Right-associative operators group from the right.  For example
\\[0.2cm]
\hspace*{1.3cm}
\texttt{a \^{ } b \^{ } c}
\\[0.2cm]
is interpreted as
\\[0.2cm]
\hspace*{1.3cm}
\texttt{a \^{ } (b \^{ } c)}).
\\[0.2cm]
Non-associative operators cannot be chained without explicit parentheses. 
The following table lists many of the common operators in \textsl{Haskell}, along with their default precedences and
associativities. (Note that these declarations can be found in the standard libraries and GHC documentation,
and some operators may have additional variants defined by specific libraries.) 
\bigskip

\begin{center}
\begin{tabular}{|l|c|l|l|}
\hline
\textbf{Operator} & \textbf{Precedence} & \textbf{Associativity} & \textbf{Description and Example} \\ \hline\hline
  function  & 10 & N/A & \texttt{f x y} applies \texttt{f} to \texttt{x} and \texttt{y}. \\
  application & & & Example: \texttt{sum [1,2,3]} \\ \hline
  \texttt{.} & 9 & right-associative  & function composition. \\
   & & & Example: \texttt{(f . g) x = f (g x)} \\ \hline
  \texttt{!!} & 9 & left-associative  & list indexing. \\
                  & & & Example: \texttt{[10,20,30] !!\;1 = 20} \\ \hline
\texttt{\^{ }} & 8 & right-associative  & power of natural numbers. \\
  & & & Example: \texttt{2 \^{ } 3 = 8} \\ \hline
  \texttt{\textasciicircum \textasciicircum } & 8 & right-associative  & floating point
                                                                                            exponentiation \\
  & & & Example: \texttt{9 \textasciicircum \textasciicircum\; (-0.5) = 3.0} \\ \hline
  \texttt{*}, \texttt{/}, & 7 & left-associative  & multiplication, floating point division \\
  & & & Example: \texttt{6 * 7 = 42} \\ \hline
  \texttt{\textasciigrave div\textasciigrave}, \texttt{\textasciigrave mod\textasciigrave} & 7 & left-associative & integer division/modulus. \\
  & & & Example: \texttt{mod 8 3 = 2} \\ \hline
  \texttt{+}, \texttt{-} & 6 & left-associative & addition and subtraction. \\ \hline
  \texttt{++} & 5 & right-associative  & List concatenation. \\
  & & & Example: \texttt{[1,2] ++ [3] = [1,2,3]} \\ \hline
  \texttt{:} & 5 & right-associative  & Cons operator for lists. \\
  & & & Example: \texttt{1 : [2,3] = [1,2,3]} \\ \hline
  \texttt{==}, \texttt{/=} & 4 & non-associative  & relational operators.  \\ \hline
  \texttt{<}, \texttt{<=}, \texttt{>}, \texttt{>=} & 4 & non-associative  & relational operators \\ \hline
  \texttt{<\$>} & 1 & left-associative  & map operator for functors \\ \hline
  \texttt{<*>} & 1 & left-associative  &  apply operator for applicatives \\ \hline
  \texttt{\&\&} & 3 & right-associative  & Boolean \textbf{and}.  \\ \hline
  \texttt{||} & 2 & right-associative  & Boolean \textbf{or} \\ \hline
  \texttt{>>=} & 1 & left-associative  & bind operator for monads \\ \hline
  \texttt{>>} & 1 & left-associative  & sequencing operator for monads \\ \hline
  \texttt{\$} & 0 & right-associative & function application operator. \\
  & & & Example: \texttt{f \$ x + y = f (x + y)} \\ \hline
\end{tabular}
\end{center}

There are many more operators in \textsl{Haskell}.  Furthermore, we can define our own operators.  This is much more
powerful  than the concept of operator overloading that we have in \textsl{Python}.  We will discuss the
details later after we have discussed the definition of functions.  Finally, we can use functions as infix
operators if we enclose them in a pair of back-quote symbols ``\texttt{\textasciigrave}''.  For example,
\texttt{div} is a function performing integer division, but instead of writing
\\[0.2cm]
\hspace*{1.3cm}
\texttt{div 8 3} \quad we can instead write \quad \texttt{8 \textasciigrave div\textasciigrave\  3}.
\\[0.2cm]
It is important to understand that function application has the highest precedence. For example, in the expression
\[
\texttt{f x + y}
\]
the function \texttt{f} is applied to \texttt{x} before adding \texttt{y}. If we want the
addition to be part of the argument to \texttt{f}, we have to use parentheses as follows: 
\[
\texttt{f (x + y)}.
\]
Alternatively, we can use the dollar-operator  and write
\[
\texttt{f \$ x + y}.
\]
This left-grouping is common for arithmetic operators and ensures consistency with standard arithmetic evaluation.

The operator \texttt{\$} is particularly useful because of its very low precedence. It allows the programmer to write expressions without a multitude of parentheses. For example, consider:
\[
\texttt{print \$ sum \$ map (\textbackslash x -> x * 2) [1,2,3]}.
\]
Without \texttt{\$}, the same expression would require nested parentheses:
\[
\texttt{print (sum (map (\textbackslash x -> x * 2) [1,2,3]))}.
\]
By declaring \texttt{\$} as having a precedence of 0 and being right associative, \textsl{Haskell} ensures that all
other operators bind more tightly, so the expression to the right of \texttt{\$} is completely grouped before
being passed as an argument. 

List operations provide a good demonstration of both precedence and associativity. The cons operator \texttt{:}
is right-associative, so 
\[
\texttt{1 :\;2 :\;3 :\;[]}
\]
is interpreted as
\[
\texttt{1 :\;(2 :\;(3 :\;[]))}.
\]
Of course, this just denotes the list $[1,2,3]$.  Th e right-associativity of the cons operator is essential
for constructing lists.

Relational operators, such as \texttt{==}, \texttt{<}, and \texttt{>=}, are declared as non-associative so that
expressions like 
\[
\texttt{a < b < c}
\]
are not allowed without parentheses. This design choice prevents ambiguous chaining of comparisons; instead,
the programmer must explicitly write 
\[
\texttt{(a < b) \&\& (b < c)}
\]
to test whether \texttt{b} lies between \texttt{a} and \texttt{c}.

\noindent
\colorbox{yellow}{\textbf{Attention:}}  There is a snag when dealing with negative numbers.  If \texttt{f} is a function that
takes one argument of type \texttt{Integer} and we want to call it with a negative number, for example with
\texttt{-42}, then we can not write the following:
\\[0.2cm]
\hspace*{1.3cm}
\texttt{f -42}
\\[0.2cm]
The reason is that \textsl{Haskell} interprets this as an expression where \texttt{42} is subtracted from
\texttt{f}.  The correct way to call \texttt{f} with an argument of \texttt{-42} is therefore to write the following:
\\[0.2cm]
\hspace*{1.3cm}
\texttt{f (-42)}


\section{Defining Functions in \textsl{Haskell}}
In \textsl{Haskell}, functions are first-class citizens and form the backbone of the language. Unlike imperative
languages where functions might be seen merely as procedures or routines, in \textsl{Haskell} every function is a \blue{pure}
mapping from inputs to outputs.  In this context, the word \blue{pure} is a technical term that means that the
function has no side effects, i.e.~it cannot change any variables or perform input or output, unless it is
specifically declared to be an \texttt{IO} function.

This section provides an in-depth exploration of function definitions in
\textsl{Haskell}. First, we discuss the basic syntax of function definitions and their type signatures. After that we discuss
\blue{matching}, \blue{guards}, \blue{higher-order functions}, \blue{currying}, \blue{lambda expressions},
recursion, and \blue{polymorphism}. 


\subsection{Introduction to Function Definitions}
At its core, a function in \textsl{Haskell} is defined by a name, a set of parameters, and an expression that computes the result. The simplest form of a function definition is:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
square :: Integer -> Integer
square x = x * x
\end{lstlisting}
Here, \texttt{square} is a function that takes a number \texttt{x} of type \texttt{Integer} and returns
\texttt{x * x}. 

Every function has a type, and while the compiler is capable of inferring types, it is a good practice to include explicit type signatures. Consider:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
add :: Integer -> Integer -> Integer
add x y = x + y
\end{lstlisting}
The type signature of \texttt{add} tells us that it takes two integers and returns an integer. Type signatures
serve as a form of documentation and help catch errors during compilation. 

\textsl{Haskell}’s type inference system can often deduce the type without explicit signatures. For instance, writing:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
multiply x y = x * y
\end{lstlisting}
allows the compiler to infer that \texttt{multiply} has a type compatible with \texttt{Num a => a -> a -> a}.
Here, \texttt{Num} is a so called \blue{type class} and the type signature
\\[0.2cm]
\hspace*{1.3cm}
\texttt{Num a => a -> a -> a}
\\[0.2cm]
tells us that if we have two arguments \texttt{x} and \texttt{y} of type \texttt{a} where the type \texttt{a}
is an instance of the type class \texttt{Num}, then the expression \texttt{multiply x y} will again have the
type \texttt{a}.  The notion of a \blue{type class} is an advanced concept that will be discussed later. 
Although type inference is possible, explicit type signatures are recommended for readability.

\subsection{Basic Function Syntax}
A function definition in \textsl{Haskell} follows the general form:
\\[0.2cm]
\hspace*{1.3cm}
$\texttt{functionName} \;\mathtt{arg}_1 \;\mathtt{arg}_2\; ... \;\mathtt{arg}_N = \mathtt{expression}$
\\[0.2cm]
Functions can have multiple parameters, and the absence of parentheses around the parameters emphasizes that
\textsl{Haskell} functions are \blue{curried} by default.  The concept of currying will be discussed now.
Consider the following example: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
add :: Int -> Int -> Int 
add x y = x + y 
\end{lstlisting}
This definition can be interpreted stating that the functions \texttt{add} takes an integer \texttt{x} and
returns a new function that itself takes an integer \texttt{y} as its argument and returns an integer.   For
clarity, we could have written the type signature of \texttt{add} as follows:
\\[0.2cm]
\hspace*{1.3cm}
\texttt{add :: Int -> (Int -> Int)}
\\[0.2cm]
This notation emphasizes that \texttt{add} takes and integer and returns a function of type \texttt{Int -> Int}.
The operator \texttt{->} is right associative and hence the types
\\[0.2cm]
\hspace*{1.3cm}
\texttt{Int -> Int -> Int} \quad and \quad \texttt{Int -> (Int -> Int)}
\\[0.2cm]
are the same.


\subsection{Pattern Matching in Function Definitions}
Pattern matching is a fundamental mechanism in \textsl{Haskell} for deconstructing data. It allows functions to perform
different computations based on the structure of their inputs. Consider the definition of the factorial
function: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
factorial :: Integer -> Integer
factorial 0 = 1
factorial n = n * factorial (n - 1)
\end{lstlisting}
Here, the pattern \texttt{0} directly matches the base case. Pattern matching can be used with more complex data types such as lists and tuples.

For example, here is a function that computes the length of a list:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
listLength :: [a] -> Int
listLength []     = 0
listLength (_:xs) = 1 + listLength xs
\end{lstlisting}
In this example, the empty list \texttt{[]} is matched by the first clause, while the pattern
\texttt{(}\textbf{\_}\texttt{:xs)}
matches any non-empty list, ignoring the head element and recursively processing the tail.  In this pattern,
the underscore \textbf{\_} denotes the so called \blue{anonymous} variable.  This is the same as in
\textsl{Python}. 

\subsection{Guards and Conditional Function Definitions}
Guards offer an alternative way to define functions that behave differently based on Boolean conditions. Instead of writing multiple equations with pattern matching, guards allow for a more readable, condition-based approach. For example, a function to compute the absolute value:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
absolute :: Integer -> Integer
absolute x 
  | x < 0     = -x
  | otherwise = x
\end{lstlisting}
Each guard (beginning with \texttt{|}) is a Boolean expression. The first guard that evaluates to \texttt{True} determines which expression is returned. The \texttt{otherwise} guard is a catch-all that always evaluates to \texttt{True}.

Guards can also be combined with pattern matching. Consider a function that classifies numbers:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
classify :: Integer -> String
classify 0 = "zero"
classify n 
  | n < 0     = "negative"
  | n > 0     = "positive"
\end{lstlisting}
This function first checks if the number is zero. If not, it uses guards to determine whether the number is negative or positive.

\subsection{Case Expressions}
A \textbf{case expression}  allows pattern matching against a value within an expression, similar to
\texttt{switch} statements in languages like \texttt{C} or \textsl{Java}. However, \textsl{Haskell}’s
\texttt{case} expressions integrate seamlessly with \textbf{pattern matching}. 
Syntactically, a \texttt{case} expression follows this general form:
\begin{verbatim}
case expression of
    pattern1 -> result1
    pattern2 -> result2
    pattern3 -> result3
    ...
\end{verbatim}
To evaluate this case expression, \textsl{Haskell} proceeds as follows:
\begin{itemize}
    \item The \textbf{expression} is evaluated.
    \item The first \textbf{pattern} that matches is chosen.
    \item The corresponding \textbf{result} is returned.
    \item If no pattern matches, a \textbf{runtime error} occurs.
\end{itemize}
Below is an implementation of a function that computes the first element of a list:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
myHead :: [a] -> a
myHead xs = case xs of
    []    -> error "No head for empty lists!"
    (x:_) -> x
\end{lstlisting}
Since \texttt{case} expressions \textbf{evaluate to values}, they can be used inline just like any other
expression.  Furthermore, \texttt{case} expressions can be combined with guards:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
describeNumber :: Int -> String
describeNumber x = case x of
    _ | x < 0     -> "Negative number"
      | x > 0     -> "Positive number"
      | otherwise -> "Zero"
\end{lstlisting}

\subsection{The Use of \texttt{where} and \texttt{let} Clauses}
Complex function definitions often benefit from local variable bindings to make the code clearer and more
modular. \textsl{Haskell} provides two constructs for this purpose: \texttt{where} clauses and \texttt{let} expressions.
\begin{enumerate}[(a)]
\item A \texttt{where} clause allows the definition of auxiliary functions and variables at the end of a function
  definition. For example, a function so solve the quadratic equation
  \\[0.2cm]
  \hspace*{1.3cm}
  $a \cdot x^2 + b \cdot x + c = 0$
  \\[0.2cm]
  can be defined as follows:
  \begin{lstlisting}[style=haskellstyle, language=Haskell]
quadratic ::  Double -> Double -> Double -> (Double, Double)
quadratic a b c = (x1, x2)
where
    discriminant = b * b - 4 * a * c
    x1 = (-b + sqrt discriminant) / (2 * a)
    x2 = (-b - sqrt discriminant) / (2 * a)
\end{lstlisting}
The \texttt{where} clause contains definitions that are local to the function \texttt{quadratic}, making the
main expression easier to read. 
\item A \texttt{let} expression provides a way to bind variables in an expression.
\begin{lstlisting}[style=haskellstyle, language=Haskell]
compute :: Int -> Int
compute x = let y = x * 2
                z = y + 3
            in z * z
\end{lstlisting}
Here, \texttt{y} and \texttt{z} are only visible in the expression following the \texttt{in} keyword.
\end{enumerate}
  A \texttt{let} expression can be used on the right hand side of a guarded equation and is then
  local to this equation, whereas the variable defined in a \texttt{where} clause are defined for all equations
  defining a function.

\subsection{Currying and Partial Application}
A unique feature of \textsl{Haskell} is that functions are curried by default. This means that every function taking multiple arguments is actually a series of functions, each taking a single argument. Consider the addition function:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
add :: Int -> Int -> Int
add x y = x + y
\end{lstlisting}
This function can be partially applied:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
increment :: Int -> Int
increment = add 1
\end{lstlisting}
Here, \texttt{increment} is a new function that adds 1 to its argument. Currying promotes code reuse and leads to elegant function composition.

\subsection{Lambda Expressions}
Lambda expressions, or anonymous functions, allow for the definition of functions without explicitly naming them. They are useful for short-lived functions, particularly when passing a function as an argument to higher-order functions. For instance:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
squares :: [Int] -> [Int]
squares xs = map (\x -> x * x) xs
\end{lstlisting}
The lambda expression \texttt{(\textbackslash x -> x * x)} takes an argument \texttt{x} and returns its square. Lambda expressions are concise and facilitate inline function definitions.

\subsection{Higher-Order Functions}

Functions that take other functions as arguments or return them as results are called higher-order functions. They are central to functional programming. For example, the \texttt{map} function applies a function to every element in a list:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
myMap :: (a -> b) -> [a] -> [b]
myMap f []     = []
myMap f (x:xs) = f x : myMap f xs
\end{lstlisting}
A custom higher-order function might filter elements in a list based on a predicate:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
myFilter :: (a -> Bool) -> [a] -> [a]
myFilter \_ [] = []
myFilter p (x:xs)
  | p x       = x : myFilter p xs
  | otherwise = myFilter p xs
\end{lstlisting}
In this definition, \texttt{myFilter} takes a predicate \texttt{p} and a list, returning a list of elements for which \texttt{p} returns \texttt{True}.

\noindent
\colorbox{yellow}{\textbf{Note}} that \textsl{Haskell} comes with the functions \texttt{map} and
\texttt{filter} that are defined exactly as we have defined the functions \texttt{myMap} and \texttt{myFilter}.
We had to rename these function when defining them ourselves because in contrast to \textsl{Python},
\textsl{Haskell} does not allow the redefinition of predefined functions.

\subsection{Recursive Function Definitions}
Since there are no control structures like \texttt{for} loops or \texttt{while} loops in \textsl{Haskell}, we
have to use recursion far more often than in \textsl{Python}.
In \textsl{Haskell}, many functions, particularly those that operate on recursive data structures such as
lists, are defined recursively.  Consider the definition of the fibonacci function that computes the $n^{\textrm{th}}$
\href{https://en.wikipedia.org/wiki/Fibonacci_sequence}{Fibonacci} number: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
fibonacci :: Integer -> Integer
fibonacci 0 = 0
fibonacci 1 = 1
fibonacci n = fibonacci (n - 1) + fibonacci (n - 2)
\end{lstlisting}
While this implementation is straightforward, it may not be efficient for large \texttt{n}. More advanced
techniques, such as \blue{memoization} or \blue{tail recursion}, can optimize recursive functions. 

Tail recursion is a form of recursion where the recursive call is the last operation in the function. Tail-recursive functions can be optimized by the compiler to iterative loops, saving stack space. For example, a tail-recursive factorial function can be written as:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
factorialTR :: Integer -> Integer
factorialTR n = factHelper n 1
  where
    factHelper 0 acc = acc
    factHelper k acc = factHelper (k - 1) (k * acc)
\end{lstlisting}
In this version, the accumulator \texttt{acc} carries the intermediate results, ensuring that the recursive call to \texttt{factHelper} is in tail position.

\subsection{Polymorphism and Overloaded Functions}

\textsl{Haskell} functions are often polymorphic, meaning that they can operate on values of various types. The function \texttt{id}, which returns its argument unchanged, is a classic example:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
id :: a -> a
id x = x
\end{lstlisting}
Here, \texttt{id} is defined for any type \texttt{a}. Polymorphism is facilitated by \textsl{Haskell}’s type system and its use of type classes, which allow functions to operate on a range of types that share common behavior.

Another example is the \texttt{const} function:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
const :: a -> b -> a
const x _ = x
\end{lstlisting}
\texttt{const} takes two arguments and returns the first, ignoring the second. Its polymorphic type signature reflects the fact that it can be applied to arguments of any types.

\subsection{The Function Composition Operator (\texttt{.})}
The function composition operator, written as a single dot ``\texttt{.}'', is one of Haskell’s most elegant and
powerful tools. It allows you to combine two functions into a new function without having to mention the
argument explicitly. In mathematical notation, function composition is written as 
\\[0.2cm]
\hspace*{1.3cm}
$(f \circ g)(x) = f\bigl(g(x)\bigr)$,
\\[0.2cm]
and in Haskell the \texttt{.} operator is defined as follows:
\begin{lstlisting}[language=Haskell]
(.) :: (b -> c) -> (a -> b) -> a -> c
f . g = \x -> f (g x)
\end{lstlisting}
The type signature \texttt{(b -> c) -> (a -> b) -> a -> c} of the function composition operator ``\texttt{.}''
tells us the following:
\begin{itemize}
  \item The operator takes two functions as arguments. The first function has type \texttt{b -> c} and the second has type \texttt{a -> b}.
  \item The composed function takes an input of type \texttt{a} and returns a result of type \texttt{c}.
\end{itemize}
In other words, if you have two functions \texttt{f :: b -> c} and \texttt{g :: a -> b}, then their composition
\texttt{(f . g)} is a function of type \texttt{a -> c} which, given an argument \texttt{x}, computes
\texttt{g x} first and then applies \texttt{f} to that result, yielding \texttt{f (g x)}. 
\vspace*{0.2cm}

\noindent
\textbf{Example}:
Consider the following functions:
\begin{lstlisting}[language=Haskell]
increment :: Int -> Int
increment x = x + 1

double :: Int -> Int
double x = x * 2
\end{lstlisting}
Using the composition operator, we can define a new function that doubles a number and then increments it:
\begin{lstlisting}[language=Haskell]
doubleThenIncrement :: Int -> Int
doubleThenIncrement = increment . double
\end{lstlisting}
Evaluating \texttt{doubleThenIncrement 3} proceeds as follows:
\\[0.2cm]
\hspace*{1.3cm}
\texttt{doubleThenIncrement 3 = increment (double 3) = increment 6 = 7}.
\\[0.2cm]
\textbf{Example}:
Another common use of function composition is to reduce parentheses in nested function calls. For instance, consider sorting a list and then reversing it:
\begin{lstlisting}[language=Haskell]
import Data.List (sort)

reverseSorted :: Ord a => [a] -> [a]
reverseSorted = reverse . sort
\end{lstlisting}
Here, \texttt{sort} is applied to the list first, and then \texttt{reverse} is applied to the sorted list. Without composition, you would have to write:
\\[0.2cm]
\hspace*{1.3cm}
\texttt{reverseSorted xs = reverse (sort xs)}.
\vspace*{0.2cm}

\noindent
The function composition operator has the following properties:
\begin{itemize}
  \item \textbf{Associativity:} Function composition is right-associative. This means that the expression
        \\[0.2cm]
        \hspace*{1.3cm}
        \texttt{f\;.\;g\;.\;h} \quad is parsed as \quad \texttt{f\;.\;(g\;.\;h)},
        \\[0.2cm]
        so you can write a chain of composed functions without excessive parentheses.
  \item By using the composition operator, we can define functions without
        mentioning their arguments explicitly. For example, instead of writing: 
        \\[0.2cm]
        \hspace*{1.3cm}
        \texttt{f x = negate (abs x)} \quad we can write \quad \texttt{f = negate\;.\;abs}.
        \\[0.2cm]
        This style is called \blue{point-free} function definition and helps to make code more concise and expressive.
\end{itemize}
The \texttt{.} operator has a very high precedence (\texttt{infixr 9}), which means it binds more tightly than most other operators. For example:
\\[0.2cm]
\hspace*{1.3cm}
\texttt{negate\;.\;abs \$ (-3)} \quad is equivalent to \quad \texttt{negate (abs (-3))}
\\[0.2cm]
Without the proper use of parentheses, the second expression would be parsed incorrectly.
\vspace*{0.2cm}

\noindent
\textbf{Summary}: The \texttt{.} operator allows you to build complex functions by composing simpler ones. Its
type signature encapsulates the idea that the output of one function becomes the input of another, and its high
precedence and associativity properties facilitate concise, point-free definitions. Whether you are chaining
arithmetic transformations or building data processing pipelines, mastering function composition is key to
writing elegant Haskell code. 
Using composition, a \blue{pipeline} of functions can be created without resorting to nested function calls. For example:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
process :: [Int] -> Int
process = sum . map square . filter even
\end{lstlisting}
Here, the list is first filtered for even numbers, then each even number is squared, and finally the squares
are summed. The composition operator makes the data flow clear and concise. 

\subsection{Point Free Style}
In point-free programming, functions are defined without explicitly mentioning their arguments. Consider the function:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
sumSquares :: [Int] -> Int
sumSquares xs = (sum . (map (\x -> x * x))) xs
\end{lstlisting}
This can be rewritten in point-free style as:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
sumSquares :: [Int] -> Int
sumSquares = sum . map (\x -> x * x)
\end{lstlisting}
Point-free style can lead to more concise definitions, though it is important to balance conciseness with
clarity.  My own experience is that it takes a while to get used to point-free style, but once you get the hang
of it, you will use it often.

\subsection{List Comprehensions in \textsl{Haskell}}
List comprehensions provide a concise and expressive way to construct lists by specifying their elements in
terms of existing lists. Inspired by mathematical set notation, list comprehensions allow you to generate new
lists by transforming and filtering elements from one or more source lists. Their elegant syntax and expressive
power make them a favorite tool for many \textsl{Haskell} programmers. 

At its core, a list comprehension has the following general syntax:
\[
[ \text{expression} \; | \; \text{qualifier}_1, \text{qualifier}_2, \dots, \text{qualifier}_n ]
\]
In this construct, the \emph{expression} is evaluated for every combination of values generated by the
qualifiers. Qualifiers can be either  \blue{generators} or \blue{filters}.  A generator has the form
\\[0.2cm]
\hspace*{1.3cm}
\texttt{pattern <- list}
\\[0.2cm]
 and is used to extract elements from an existing list, while a filter is simply a
Boolean expression that restricts which elements are included. 
We begin with  a simple example where we generate a list of squares for the numbers from 1 to 10:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
squares :: [Int]
squares = [ x * x | x <- [1..10] ]
\end{lstlisting}
Here, \texttt{x <- [1..10]} is a generator that iterates over the numbers 1 through 10, and the expression
\texttt{x * x} calculates the square of each number. 

List comprehensions also allow you to filter elements by adding a Boolean condition. For instance, to generate
a list of even squares from 1 to 10 we can write the following:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
evenSquares :: [Int]
evenSquares = [ x * x | x <- [1..10], even x ]
\end{lstlisting}
In this example, the qualifier \texttt{even x} acts as a filter, ensuring that only even values of \texttt{x}
are considered. As a result, only the squares of even numbers are produced. 

List comprehensions can also combine multiple generators to produce lists based on the Cartesian product of
several lists. For example, the following comprehension generates pairs of numbers where the first element is
taken from \texttt{[1,2,3]} and the second from \texttt{[4,5]}: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
pairs :: [(Int, Int)]
pairs = [ (x, y) | x <- [1,2,3], y <- [4,5] ]
\end{lstlisting}
This comprehension evaluates the tuple \texttt{(x, y)} for each combination of \texttt{x} and \texttt{y}, resulting in a list of pairs.

\subsection{Haskell Sections}
Haskell sections are a syntactic convenience that allow for the partial application of binary
operators. In essence, a section is an expression in which one operand of a binary operator is
fixed, yielding a new function that awaits the missing operand. For example, the section
\texttt{(+1)} represents a function that adds 1 to its argument and is therefore equivalent to the lambda expression
$\lambda x \to x + 1$. Likewise, \texttt{(1+)} is equivalent to $\lambda x \to 1 + x$. This
mechanism enables concise function definitions without the need for explicit lambda notation. 

The general form of a section is either:
\\[0.2cm]
\hspace*{1.3cm}
\texttt{(o e)} \quad or \quad \texttt{(e o)}
\\[0.2cm]
where \texttt{o} is a binary operator and \texttt{e} is an expression that serves as one of its
argument.  The missing argument is supplied when the section is applied to an argument.  Therefore,
\begin{enumerate}
\item \texttt{(o e)} \quad is equivalent to \quad \texttt{\symbol{92}x -> x o e}, and
\item \texttt{(e o)} \quad is equivalent to \quad \texttt{\symbol{92}x -> e o x}.
\end{enumerate}
For example, 
\\[0.2cm]
\hspace*{1.3cm}
\texttt{(+ 1)} \quad is equivalent to \quad \texttt{\symbol{92}x -> x + 1}.
\\[0.2cm]
Hence we can define a function that increments its argument as follows:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
increment :: Num a => a -> a
increment = (+1)
\end{lstlisting}

Sections work for any infix operator, and their behavior is influenced by the operator’s
properties. Consider the subtraction operator: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
subtractFromTen :: Num a => a -> a
subtractFromTen = (10-)
\end{lstlisting}
Here, \texttt{(10-)} is a section that subtracts its argument from 10.  Notice that the order
matters: while \texttt{(10-)} computes $10 - x$, the expression \texttt{(-10)} is simply a negative
literal representing $-10$.  Let us take a look at a few more examples:
\begin{itemize}
  \item \texttt{(==0)} is a section that checks if its argument is equal to 0:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
isZero :: (Eq a, Num a) => a -> Bool
isZero = (==0)
\end{lstlisting}
  \item \texttt{(/2)} divides its argument by 2:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
half :: Fractional a => a -> a
half = (/2)
\end{lstlisting}
  \item \texttt{(2\^{ })} computes 2 raised to the power of its argument:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
powerOfTwo :: (Integral b, Num a) => b -> a
powerOfTwo = (2^)
\end{lstlisting}
\end{itemize}

\section{Algebraic Data Types in Haskell}
In Haskell, an \textbf{Algebraic Data Type (\textsc{Adt})} is a way to define new types by combining other types. \textsc{Adt}s
allow developers to construct complex data structures in a type-safe and declarative manner. The term
"algebraic" comes from the fact that these data types are constructed using algebraic operations such as sums
(alternatives) and products (combinations). 
\textsc{Adt}s are primarily classified into two types:
\begin{itemize}
    \item \textbf{Sum types} (disjoint unions), which represent a choice between multiple alternatives.
    \item \textbf{Product types}, which combine multiple values together.
\end{itemize}

\subsection{Defining Algebraic Data Types}
\textsc{Adt}s are defined in Haskell using the \texttt{data} keyword. The general syntax is:

\begin{lstlisting}[style=haskellstyle, language=Haskell]
  data TypeName = Constructor1 Type1 Type2 ...
                | Constructor2 Type3 Type4 ...
                ...
                | Constructor7 Type8 Type9 ...
\end{lstlisting}

Each constructor represents a possible form that a value of this type can take.

\subsection{Example: Defining a Simple \textsc{Adt}}
Consider defining a type to represent a traffic light:

\begin{lstlisting}[style=haskellstyle, language=Haskell]
data TrafficLight = Red | Yellow | Green
\end{lstlisting}
This defines \texttt{TrafficLight} as a type with three possible values: \texttt{Red}, \texttt{Yellow}, and
\texttt{Green}. Each value is represented by a constructor with no associated data. 

\subsection{Sum Types (Disjoint Unions)}
A sum type is a type that can take on one of several different values. The \texttt{|} symbol is used to define different constructors. For example:

\begin{lstlisting}[style=haskellstyle, language=Haskell]
data Shape = Circle Float | Rectangle Float Float
\end{lstlisting}
Here, a \texttt{Shape} can either be a \texttt{Circle} (with a radius of type \texttt{Float}) or a
\texttt{Rectangle} (with width and height of type \texttt{Float}). 

\subsection{Product Types}
Product types combine multiple values together in a single constructor. A classic example is a coordinate point:

\begin{lstlisting}[style=haskellstyle, language=Haskell]
data Point = Point Float Float
\end{lstlisting}
This defines a \texttt{Point} type where each \texttt{Point} consists of two \texttt{Float} values representing
the x and y coordinates. 

\subsection{Pattern Matching with \textsc{ADT}s}
Pattern matching allows functions to be defined by destructuring \textsc{Adt} values. Here is an example using the \texttt{Shape} type:

\begin{lstlisting}[style=haskellstyle, language=Haskell]
area :: Shape -> Float
area (Circle r) = pi * r * r
area (Rectangle w h) = w * h
\end{lstlisting}
The function \texttt{area} takes a \texttt{Shape} and returns its area by matching on its constructor.

\subsection{Recursive Data Types}
\textsc{Adt}s can also be recursive, allowing the definition of complex data structures such as lists or trees. Consider the definition of a binary tree:

\begin{lstlisting}[style=haskellstyle, language=Haskell]
data Tree a = Leaf a | Node (Tree a) (Tree a)
\end{lstlisting}
This defines a tree where each node contains either a single value (\texttt{Leaf}) or two subtrees
(\texttt{Node}). 

\section{Type Classes in Haskell}
A \textbf{type class} in Haskell is a mechanism for defining a set of functions that can operate on multiple
types. It enables \textit{ad-hoc polymorphism}, meaning that a single function can have different
implementations depending on the type of its arguments. Type classes are similar to interfaces in
object-oriented languages but are more flexible because they are based on Haskell’s strong type system. 

A type class is defined using the \texttt{class} keyword. Here is a simple example:

\begin{lstlisting}[style=haskellstyle, language=Haskell]
class Eq a where
    (==) :: a -> a -> Bool
    (/=) :: a -> a -> Bool
\end{lstlisting}

This declaration defines an \texttt{Eq} type class with two methods: \texttt{(==)} for equality and \texttt{(/=)} for inequality. Any type that is an instance of \texttt{Eq} must provide implementations for these functions.

\subsection{The \texttt{Eq} Type Class}
The \texttt{Eq} type class is used for types that support equality comparison. It is defined in the Haskell Prelude as:

\begin{lstlisting}[style=haskellstyle, language=Haskell]
class Eq a where
    (==) :: a -> a -> Bool
    (/=) :: a -> a -> Bool
    x /= y = not (x == y)
    x == y = not (x /= y)
\end{lstlisting}

The last two lines provide default implementations: if \texttt{(==)} is defined, then \texttt{(/=)} can be
derived automatically and vice versa. 

All standard Haskell types, such as \texttt{Int}, \texttt{Char}, and \texttt{Bool}, are instances of
\texttt{Eq}. User-defined types can also be made instances of \texttt{Eq}: 

\begin{lstlisting}[style=haskellstyle, language=Haskell]
data Color = Red | Green | Blue

instance Eq Color where
    Red   == Red   = True
    Green == Green = True
    Blue  == Blue  = True
    _     == _     = False
\end{lstlisting}
Alternatively, Haskell allows automatic instantiation of the type class \texttt{Eq} for algebraic data types
via the keyword \blue{\texttt{deriving}}: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
data Color = Red | Green | Blue deriving Eq
\end{lstlisting}

\subsection{The \texttt{Ord} Type Class}
The \texttt{Ord} type class is used for totally ordered data types. It extends \texttt{Eq} and defines methods
for comparison: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
class Eq a => Ord a where
    compare :: a -> a -> Ordering
    (<), (<=), (>=), (>) :: a -> a -> Bool
    max, min :: a -> a -> a
\end{lstlisting}
Here, the function \texttt{compare} returns an \texttt{Ordering} value, which can be \texttt{LT} (less than),
\texttt{GT} (greater than), or \texttt{EQ} (equal).  The other comparison functions can then be defined in
terms of \texttt{compare}: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
compare x y
    | x == y    = EQ
    | x <= y    = LT
    | otherwise = GT
\end{lstlisting}
Like \texttt{Eq}, \texttt{Ord} instances can be derived automatically for algebraic data types:
\begin{lstlisting}[style=haskellstyle, language=Haskell]
data Color = Red | Green | Blue deriving (Eq, Ord)
\end{lstlisting}
For derived \texttt{Ord} instances, the order of constructors in the data declaration determines the comparison
result. For example: 
\begin{lstlisting}[style=haskellstyle, language=Haskell]
Red   < Green == True
Green > Blue  == False
\end{lstlisting}

\subsection{Custom Instances of \texttt{Ord}}
If a type does not have a natural ordering, we can define our own:

\begin{lstlisting}[style=haskellstyle, language=Haskell]
data Size = Small | Medium | Large

instance Eq Size where
    Small  == Small  = True
    Medium == Medium = True
    Large  == Large  = True
    _      == _      = False

instance Ord Size where
    Small  <= _      = True
    Medium <= Large  = True
    Medium <= Medium = True
    Large  <= Large  = True
    _      <= _      = False
\end{lstlisting}
With this instance, Haskell can compare \texttt{Size} values, allowing operations like sorting.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "haskell"
%%% End:
